LDpred2: better, faster, stronger

### 2020-04-24 paper/paper-ldpred2.tex
save

\begin{figure}[h]
	\centerline{\includegraphics[width=0.85\textwidth]{QC-CAD}}
	\caption{Results of the real data applications using published external summary statistics.
		Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
		%See corresponding values in table S2.
	}
	\label{fig:QC-CAD}
\end{figure}
[TALK ABOUT QUALITY CONTROL ON AF?]

@@

@@

For one marginal effect $\hat{\gamma}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
Then,
$\text{var}(\hat{\gamma}) = \dfrac{(y^* - \hat{\gamma} g^*)^T (y^* - \hat{\gamma} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{sd}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{sd}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.

Let us go back to the formula we derived before: $\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}}$. 
Note that $\gamma$ are the effects on the allele scale while we denote by $\beta = S \gamma$ the effects of the scaled genotypes.
As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$. Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{sd}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred. Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\gamma}) ~ \sqrt{N}\right)$.
Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.

@@

To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.

First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
\begin{equation}\label{eq:beta_res}
\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{K}_{\text{-}j, j}
\end{equation}
where $\boldsymbol{K}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix and $\boldsymbol{\hat{\beta}}$ are the marginal effect sizes.
Then, they compute the probability that variant $j$ is causal as 
\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
which we rewrite as
\begin{equation}
\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.\label{eq:postp}
\end{equation}
Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.

@@

%\begin{minipage}{.7\textwidth}
\begin{algorithm}[H]
\caption{LDpred, with hyper-parameters $p$ and $h^2$}\label{algo:ldpred}
\begin{algorithmic}[1]
	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
	\For {each variant $j$} \Comment{All variants}
	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
	\EndFor
	\State Compute $\boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
	\If {$k > N_\text{burn-in}$} 
	\State Increment $\boldsymbol{\Omega}$ with $\boldsymbol{\omega}$ 
	\EndIf
	\EndFor
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$
	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
\end{algorithmic} 
\end{algorithm}
%\end{minipage}

[CHANGE NOTATION OF SUMSTATS / SD -> SE?]

@@

We recommend to run LDpred2 per chromosome.
Even if it is possible to run LDpred2 genome-wide, this approach has two limitations. First, it can be memory and computationally demanding to do so. For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory. 
Doubling to 2M variants would require 128 GB of RAM to store the matrix.
Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
For example, autoimmune diseases clearly show larger effects on chromosome 6 compared to other chromosomes.
Assuming the same model for genetic effects genome-wide could result in severe model misspecification, which would lead to suboptimal predictive performance.
Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.

>>>>

\begin{figure}[h]
	\centerline{\includegraphics[width=0.85\textwidth]{QC-CAD}}
	\caption{Results of the real data applications using published external summary statistics.
		Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
		%See corresponding values in table S2.
	}
	\label{fig:QC-CAD}
\end{figure}

[TALK ABOUT QUALITY CONTROL ON AF?]

@@

[NEED TO UNDERSTAND WHY AUTO DOES NOT WORK FOR SOME TRAITS -> CAD BECAUSE DIFFERENT N // T1D PROBABLY BECAUSE OF SMALL SAMPLE SIZE // ASTHMA? AND PRCA?]

[TALK ABOUT MODEL FOR BINARY DATA]

@@

For one marginal effect $\hat{\gamma}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
Then,
$\text{var}(\hat{\gamma}) = \dfrac{(y^* - \hat{\gamma} g^*)^T (y^* - \hat{\gamma} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{se}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{se}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.

Let us go back to the formula we derived before: $\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}}$. 
Note that $\gamma$ are the effects on the allele scale while we denote by $\beta = S \gamma$ the effects of the scaled genotypes.
As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$. Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{se}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred. Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{se}(\hat{\gamma}) ~ \sqrt{N}\right)$.
Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.

@@

To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.
First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
\begin{equation}\label{eq:beta_res}
\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{K}_{\text{-}j, j}
\end{equation}
where $\boldsymbol{K}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix and $\boldsymbol{\hat{\beta}}$ are the marginal effect sizes.
Then, they compute the probability that variant $j$ is causal as 
\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
which we rewrite as
\begin{equation}
\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.\label{eq:postp}
\end{equation}
Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.

@@

%\begin{minipage}{.7\textwidth}
\begin{algorithm}[H]
\caption{LDpred, with hyper-parameters $p$ and $h^2$}\label{algo:ldpred}
\begin{algorithmic}[1]
	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
	\For {each variant $j$} \Comment{All variants}
	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
	\EndFor
	\State Compute $\boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
	\If {$k > N_\text{burn-in}$} 
	\State Increment $\boldsymbol{\Omega}$ with $\boldsymbol{\omega}$ 
	\EndIf
	\EndFor
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$ \Comment{Average of all $\boldsymbol{\omega}$ after burn-in}
	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
\end{algorithmic} 
\end{algorithm}
%\end{minipage}

[CHANGE NOTATION OF SUMSTATS?]

@@

We recommend to run LDpred2 per chromosome.
Even if it is possible to run LDpred2 genome-wide, this approach has two limitations. First, it can be memory and computationally demanding to do so. For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory. 
Doubling to 2M variants would require 128 GB of RAM to store the matrix.
Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
Assuming the same model for genetic effects genome-wide could result in severe model misspecification, which would lead to suboptimal predictive performance.
Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.
